# Data Importing and Exporting 
## Introduction to Data Importing 
Explanation of different data formats (e.g., CSV, Excel, JSON).
Overview of R functions for reading data (read.csv(), read_excel(), read.table(), etc.).

Loading data in R can be very stressful since every file format has a function to import the data. However, it can be very simple when wel explained. Here are some of the most commonly used data file formats R and their importation;-

- Comma Separated Value(CSV - .csv) files imported by `read.csv("filepath")` function
- Excel(.xlsx) files loaded by `read_excel("filepath")` function`
- XML(.xml) imported by `read.xml("filepath")` function` 
- Javascript Object Notation (JSON) by  `read.json("filepath")``

There are more data files with different formats that can be used in R, their importation will be explained later in the course. 

## Demonstration of Data Importing
- **Importing a CSV file in R**

There are different formats to import a csv file in R. We will use the `read.csv()` and `read.delim()` which are functions from the baseR. 

Reading a csv file using `read.csv()` function. 
```{r}
# read the csv file to the data frame 
df <- read.csv('data/hotel_bookings_clean.csv')
head(df) # show first few rows 
```

Now, lets read the same csv file with `read.delim()` function. Remember to add "," as the separator. `sep=","`. 
```{r}
# Read the csv file 
df <- read.delim("data/hotel_bookings_clean.csv", sep = ",")
head(df) # show the first few rows
```
Lets also read the csv file using the `read.table()` function(note the `sep=","` argument);
```{r}
df <- read.table("data/hotel_bookings_clean.csv", sep = ",")
head(df)
```

The `read_csv()` function is one more way to read a csv file. This function comes from the `readr` package that is installed by;
```
install.packages("readr")
```

and imported by;
```{r}
library(readr)
```

Lets read a csv file using the `read_csv()`(note the '_' instead of a '.') from the `readr` library. 
```{r}
df <- read_csv("data/hotel_bookings_clean.csv", show_col_types = FALSE)
head(df)
```

`read.csv()`, `read.delim()`, `read.table()` and `read_csv()` are the functions used to read CSV files in R. 

- **Importing Excel file in R**

In this section we will read the HR employee data for descriptive and inferential analytics. We will use the `read_excel` from the `readxl` package. It can be imported by;
```{r}
library(readxl)
```

Let's load the excel data into R data frame. 
```{r}
df <- read_excel("data/HR_Employee_Data.xlsx")
head(df)
```

If they are multiple worksheets in the excel, the read_excel has an argument `sheet` for sheet number(By default, sheet 1 is loaded. Lets load the first sheet. 
```{r}
df_sheet1 <- read_excel("data/HR_Employee_Data.xlsx", sheet=1)
head(df_sheet1)
```

A detail documentation of the `read_excel()` function can be found [here](https://www.rdocumentation.org/packages/readxl/versions/1.4.1/topics/read_excel)

<- **Add more methods**->

## Introduction to Data Exporting
